Development of a Finger-Key Identification Module for a Touch Typing Trainer

### 2022-06-23 journal-article/journal-article.tex
journal-article: finish journal article

\documentclass{report}
%%% LT-SKIP-BEGIN

% Page Size
\usepackage[letterpaper, portrait, margin=1in, left=1.5in]{geometry}

% Spacing
\usepackage{setspace}
\setlength{\parskip}{\baselineskip}
\doublespacing{}

% Bibliography
\usepackage[american]{babel}
\usepackage[style=apa, citestyle=apa, sortcites=true, backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\usepackage[babel,threshold=2]{csquotes}

% Graphics
\usepackage{graphicx}
\graphicspath{ {./images/} }

@@

@@

\DeclareAcronym{roi}{
	short=ROI,
	long=Region of Interest
}

@@

% Chapter Formatting
\usepackage{titlesec}
\titleformat{\chapter}[display]
{\centering\LARGE\normalfont\bfseries}
{Chapter \thechapter}
{0em}
{\vspace{-1.5ex}}

% Misc
\usepackage{gensymb}
\usepackage{amsmath}

% Graphs
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{external}
\pgfplotsset{width=1\textwidth,compat=1.9}

% Tables
\usepackage{booktabs}
\usepackage{supertabular}

% Adjust Figure Spacing
\setlength\intextsep{1cm}

% Add Gantt Chart
\usepackage{pdfpages}
\usepackage{pdflscape}

% Center table with controlled width
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

% List of Equations
\usepackage[titles]{tocloft}
\newcommand{\listequationsname}{List of Equations}
\newlistof{equations}{equ}{\listequationsname}
\newcommand{\labelequations}[1]{%
	\addcontentsline{equ}{equations}{\numberline{\theequation}#1}\par}
\setlength{\cftequationsindent}{1.5em}
\setlength{\cftequationsnumwidth}{2.3em}

\addto\captionsamerican{
	\renewcommand{\contentsname}
	{Table of Contents}
}

% Subfigure
\usepackage{subcaption}

%%% LT-SKIP-END

@@

\pagenumbering{gobble}
\begin{titlepage}
	\centering

	\hspace{0pt}
	\vfill

	\includegraphics[width=0.2\textwidth]{upc.png}
	\includegraphics[width=0.2\textwidth]{dcs.png}
	\par\vspace{1cm}

	\textsc{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}
	\par\vspace{0.5cm}

	\hrulefill{}
	\par\vspace{0.25cm}
	A Special Project Presented to the\\
	Faculty of the Department of Computer Science,\\
	University of the Philippines Cebu

	\par\vspace{0.25cm}
	In Partial Fulfillment\\
	Of the Requirements for the Degree\\
	Bachelor of Science in Computer Science\\
	\par\vspace{0.25cm}
	\hrulefill{}
	\par\vspace{0.5cm}

	Oscar Vian L. Valles\\
	BS Computer Science
	\par\vspace{0.5cm}

	Dhong Fhel K. Gom-os\\
	Adviser
	\par\vspace{0.5cm}

	June 2022
	\vfill
	\hspace{0pt}
\end{titlepage}

\begin{center}

	\hspace{0pt}
	\vfill

	\includegraphics[width=0.2\textwidth]{upc.png}
	\includegraphics[width=0.2\textwidth]{dcs.png}
	\par\vspace{0.25cm}

	\textbf{\uppercase{University of the Philippines Cebu}}\\
	College of Science\\
	Department of Computer Science

	\par\vspace{1cm}

	\textbf{\uppercase{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}}

	\par\vspace{1cm}

	\textbf{Permission is given for the following people to have access to this thesis:}

	\par\vspace{0.25cm}

	\begin{tabular}{ | P{0.8\textwidth} | P{0.1\textwidth} | }
		\hline
		Available to the general public                                 & Yes \\
		\hline
		Available only after consultation with author or thesis adviser & Yes \\
		\hline
		Available to those bound by confidentiality agreement           & Yes \\
		\hline
	\end{tabular}
	\par\vspace{1cm}

	\line(1,0){0.6\textwidth}\\
	Oscar Vian L. Valles\\
	Student
	\par\vspace{0.5cm}

	\line(1,0){0.6\textwidth}\\
	Dhong Fhel K. Gom-os\\
	Special Problem Adviser

	\vfill
	\hspace{0pt}

\end{center}

\begin{center}
	\uppercase{\textbf{APPROVAL SHEET}}
\end{center}

\par\vspace{0.5cm}
\noindent
The faculty of the University of the Philippines Cebu, College of Science,
Department of Computer Science approves this Special Problem entitled:
\par\vspace{0.5cm}

\begin{center}
	\textbf{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}\\
	by\\
	Oscar Vian L. Valles

	\par\vspace{0.5cm}

	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Dhong Fhel K. Gom-os\\
		Special Problem Adviser

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date Signed
	\end{minipage}
	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Aileen Joan O. Vicente\\
		Chair, Department of Computer Science

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date Signed
	\end{minipage}

	\par\vspace{2cm}

	\line(1,0){0.405\textwidth}\\
	Prof. Nelia S. Ereno\\
	Dean, College of Science

	\par\vspace{0.5cm}
	\line(1,0){0.405\textwidth}\\
	Date Signed

	\par\vspace{2cm}
\end{center}

\pagenumbering{roman}

\addcontentsline{toc}{chapter}{Abstract}
\begin{center}
	\LARGE\textbf{Acknowledgment}
\end{center}

I would like to express my deepest appreciation for the guidance that my Special
Project Adviser, Dhong Fhel K. Gom-os, has given me throughout the duration of
the project. His technical knowledge has been a key factor to the success of the
finger-key identification module. The paper would not have been possible without
his unwavering support and commitment.

I also greatly appreciate the advice that Prof. Aileen Joan O. Vicente and Ms.
Sram J Isle L. Santillan have given me. The constructive criticism that I
received was instrumental in giving me a clear direction on where I want to take
this research.

I am particularly grateful to Jack Bartnik, the developer of Monkeytype. The
application that they developed was the main source of inspiration for this
project. I would also like to thank Jack for their generosity in sharing
statistics concerning the typing speed of the users of Monkeytype.

I would like to acknowledge that this research was partially supported by the
Department of Science and Technology --- Science Education Institute.

I give my heartfelt thanks to Ana Janelle H. Ca\~{n}ete for her love and comfort
in these trying times.

To friends, family, and other significant people in my life, thank you for
supporting me thus far.

\newpage
\addcontentsline{toc}{chapter}{Abstract}
\begin{center}
	\LARGE\textbf{Abstract}
\end{center}

Typing tests list out words for the user to repeat. These tests measure metrics
that quantify the user's ability to type. However, conventional typing metrics
do not measure correct finger placement. This aspect of typing may affect the
user's health. It is also a crucial aspect of keyboard typing education. As
such, a method to identify which finger is used to press which key is
beneficial. This paper introduces a technique to achieve this by developing a
finger-key identification module that utilizes computer vision algorithms to
detect the keys in a keyboard, and a ready-made machine learning solution to
track fingers while typing. This module was successful in finger-key
identification with an accuracy of 99.58\% in a dataset of 942 keypresses. The
module also had an average finger-key identification time of 0.083 seconds,
which allows the module to perform real-time finger-key identification for
typing speeds up to 143.030 Words per Minute which is well beyond the median
typing speed of the population. Furthermore, two metrics for correct finger
placement were defined and were implemented in a proof-of-concept trainer: (1)
Finger Placement Accuracy which computes the number of keys pressed using the
correct finger in a typing test sequence, and (2) Historical Finger Placement
Accuracy which evaluates the user's ability to press a certain key with the
correct finger over multiple typing test sequences.

\addcontentsline{toc}{chapter}{Acronyms}
\printacronyms{}
\newpage

\addcontentsline{toc}{chapter}{List of Tables}
\listoftables
\newpage

\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures
\newpage

\addcontentsline{toc}{chapter}{List of Equations}
\listofequations

\tableofcontents
\newpage

\pagenumbering{arabic}

\chapter{Introduction}

\section{Background of the Study}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests
list out words that are then typed out. The entered keys are then compared to
check if the user has typed the expected letter. At the end of the test, the
time taken is calculated, and certain metrics are given. These metrics include
\ac{wpm} and Accuracy \parencite{arif2009}. However, this method of examination
leaves out a crucial part of typing --— correct finger placement.

Incorrect finger placement may affect hand and finger health. Incorrect
placement of the fingers may cause these hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \parencite{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions, such as in typing, may cause injuries such
as \ac{cts} \parencite{toosi2015}.

Correct finger placement is usually taught at the beginning using a diagram,
with each key being associated with a specific finger. For instance, the letter
Q in a QWERTY layout should be pressed using the fifth digit, or the little
finger, of the left hand, and this is shown on instruction manuals by coloring
the fifth digit and the key Q with the same color or by placing the letters
directly on the fingers \parencite{dobson2009touch}.

Furthermore, touch typing using correct finger placement is frequently taught in
the beginner level \parencite{donica2018}. This means that there is a need to
weed out bad habits that may develop, like using the index finger for pressing
the spacebar or backspace. However, it is impractical for an educator to check
each student if they are not performing these movements as these may only show
for a small period and may not be caught in time.

Thus, there is a need for automatically identifying which finger is used to
press which key during typing --- hereby defined as finger-key identification.

\section{Research Questions}
\begin{enumerate}
	\item What algorithm or technique using computer vision is capable of
	      finger-key identification.
	\item What is the setup and configuration of a single optical camera can be
	      used to capture images during keyboard typing for real-time finger-key
	      identification using the algorithm identified in 1?
	\item How to design and develop a finger-key identification module using
	      the algorithm in 1 and camera setup and configuration in 2?
	\item How accurate is the module developed in 3 in finger-key identification?
	\item Is the module developed in 3 fast enough to run in real-time?
	\item What new keyboard typing metric to develop that considers both the key and
	      finger used to press as parameters?
\end{enumerate}

\section{Research Objectives}

\subsection{General Objectives}
To create a finger-key identification module that accurately identifies what
finger was used to press a key at a point in time to help develop better typing
habits and healthier typing ergonomics.

\subsection{Specific Objectives}
\begin{enumerate}
	\item Develop an algorithm or technique using computer vision capable of
	      finger-key identification.
	\item Identify the setup and configuration of a single optical camera that can
	      be used to capture images during keyboard typing for finger-key
	      identification using the algorithm identified in 1
	\item Design and develop a module using the algorithm in 1 and camera setup
	      and configuration in 2
	\item Determine the accuracy of the module developed in 3 in finger-key
	      identification
	\item Determine if the module developed in 3 is fast enough to run in
	      real-time
	\item Develop a new keyboard typing metric that uses both the key and
	      finger used to press as parameters
\end{enumerate}

\section{Scope and Limitation}
The experimental setup was simplified and tightly controlled. It was not
representative of real-world conditions. The colors of the keyboard, cables, and
surface were all controlled. So were the lighting and camera used.

The supported keyboard of the module was limited to one specific type of
keyboard. A 60\% Keyboard in a slightly modified \ac{ansi} layout with
light-colored keys was used as shown in Figure~\ref{fig:60}. This type of
keyboard only has the alphanumeric portion of a full-size keyboard. This limited
the number of keys to be checked and reduced the overall movement of the hand
during typing. The keyboard layout was a slightly modified \ac{ansi} layout. The
majority of the layout still followed the base \ac{ansi} layout, with the
alphanumeric portion remaining exactly the same.

The module's effect on typing was not tested. This includes the module's effect
on typing in education and typing in health, as the focus of the research was on
the development and testing of the finger-key identification module only.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth]{60.png}
	\caption{A 60\% keyboard in modified \ac{ansi} layout.}
	\label{fig:60}
	\centering
\end{figure}

\section{Significance of the Research}
This research is beneficial for all users of physical keyboards. These include a
majority of the population as there are a lot of professions that rely heavily
on keyboards. Examples include developers, physicians, educators, and
accountants. By having better ergonomics while typing --- by touch typing
correctly --- wrist injuries can be prevented, and typing speed may be
increased.

This research also helps educators, especially early educators teaching beginner
typists. By automatically checking for correct technique, the burden of checking
each student is reduced, and directed interventions for bad habits can be easily
created as students with these bad habits are easily identified.

This research has a direct impact on people that have hand or wrist injuries
that are caused by poor typing habits. By correcting these poor habits, pain
from these injuries will be reduced, and even be prevented from occurring in the
first place. A specific example of this is by reducing ulnar deviation which
affects the nerve that is indicative of \ac{cts} \parencite{toosi2015}.

\chapter{Review of Related Literature}

\section{Keyboard Typing}

Keyboard typing is the process of using a keyboard to input characters in a
system. In the context of this paper, keyboard typing will refer to the act of
using a physical keyboard to input characters into a computer system.

\subsection{Keyboard Layouts and Form Factors}

One key characteristic of a keyboard is its physical attributes. Keyboards come
in a lot of layouts and form factors. Keyboard layouts are the shapes, size, and
positions of a key on a keyboard while the form factor of a keyboard refers to
its shape and dimensions. The form factor also refers to the number of keys included
in the keyboard \parencite{parkkinen2018}. By combining different layouts and
form factors, different permutations of a keyboard can be created.

Different keyboard layouts and form factors also produce different effects for
the user. This is due to how vastly different some keyboard layouts and form
factors are from one another. Some layouts focus on ergonomics, while others
focus on typing speed. Some form factors were designed for aesthetics, while
others focus on comfort and health. As such, different layouts may affect typing
performance, ergonomics, and long-term health effects \parencite{ciobanu2015}.

\subsubsection{ANSI and ISO Layout}

There are two common keyboard layouts around the world --- \ac{iso} and \ac{ansi}.

\citeauthor{ansi} is the standard that first defined the \ac{ansi} layout.
Figure~\ref{fig:ansi} illustrates what the \ac{ansi} layout looks like. This
layout is also used by countries other than America. Examples of countries that
use this layout as its standard are the Philippines, China, and Korea
\parencite{apple-layout}. However, these countries also opt to modify the layout
by adding extra layers to accommodate other character sets.

\citeauthor{iso} is the standard series that defines a framework that is used to
create other layouts. Layouts created from this standard are colloquially called
\ac{iso} Layouts. Countries around the world use this framework to create
layouts that fit the characters in their language. Examples of countries that
use this framework to create their layout are France, Greece, Canada, and Sweden
\parencite{apple-layout}.

Both of these layouts usually utilize the same key ordering. This ordering is
commonly called QWERTY, based on the first five characters of the first row
of this specific layout.

There are other layouts available, however, they are not as common as the two
previously mentioned layouts. Examples of this include \ac{jis}. Other esoteric
layouts, like Tsangan or split-backspace, also exist. These layouts modify the
\ac{iso} and \ac{ansi} standards by adding or removing certain keys to fit the
character set of a language, or for additional keys. Other layouts are also the
same as \ac{ansi} or \ac{iso}, however, these layouts change the arrangement of
the alphabet within the keyboard.

Despite the ubiquity of these common layouts, studies have shown that these
layouts are not ergonomic. The main issue with these layouts is the random
configurations of the letters. The randomness of the layout necessitates
memorization of the layout which reduces the ease of learning, reduces performance
in typing by reducing speed, and increases typing errors
\parencite{ciobanu2015}.

\subsubsection{Keyboard Form Factors}

There is only one common keyboard form factor used worldwide: the full-size
keyboard. This keyboard contains all the keys specified in the keyboard
layout. This includes the alphanumeric keys, the function keys, the navigation
cluster, and the Numpad.

Other common keyboard form factors are based on the full-size keyboard. The name
of these layouts, 60\%, 75\%, and 80\% reference the remaining number of keys
after cutting a portion off from the full-size keyboard. The 60\% keyboards only
contain the alphanumeric cluster while the 80\% and 75\% layouts retain the
navigation cluster and the function keys \parencite{parkkinen2018}. The main
draw for using keyboards with reduced sizes is for aesthetics, space
constraints, and ergonomics.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{ansi.png}
	\caption{\ac{ansi} Keyboard layout with form factors. Reprinted from \fullcite{figure-ansi}}
	\label{fig:ansi}
	\centering
\end{figure}

\subsubsection{Ergonomic Keyboards Layouts and Form Factors}

There have been other keyboard layouts and form factors created to mitigate
common issues associated with QWERTY layouts. These include Colemak, Dvorak, and
Alphabetical layouts. However, studies have shown that the layout itself does
not matter as beginners do not necessarily see the keyboard as a structured set,
but rather as a random collection of characters, even if it is alphabetized
\parencite{norman1982},

A different form factor has a great effect on ergonomics. One such example of a
form factor is an ergonomic keyboard developed by Microsoft called Microsoft
Natural MultiMedia Keyboard. \citeauthor{ripat2010} used this keyboard in
determining that ergonomic keyboards can help in reducing symptoms of \ac{neck}.
Figure~\ref{fig:mn} shows the layout of the Microsoft Natural MultiMedia
Keyboard.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{mn.png}
	\caption{Microsoft Natural MultiMedia Keyboard. Reprinted from \fullcite{mn}}
	\label{fig:mn}
	\centering
\end{figure}

There are other form factors other than the full-size keyboard and variations
thereof that focus on ergonomics. One such example is a split keyboard layout
where the keyboard is split in half, one for the left hand and one for the
right. One such benefit, according to \citeauthor{ergodox}, is a more relaxed
position due to typing at shoulder width.

\subsection{Keyboard Typing Metrics}

There are numerous metrics used to quantify keyboard typing performance. Two
common metrics used in the majority of typing tests include Accuracy and Speed.

\subsubsection{Standardized Keyboard Typing Assessments}

To be able to measure these metrics, a keyboard typing assessment needs to be
done. However, there are no standardized keyboard typing assessments
\parencite{donica2018}. As such, teaching methods and assessments, like
Keyboarding without Tears, Monkeytype, and Keybr, may produce different metrics
for the same typist due to their difference in conducting the assessment.

\subsubsection{Speed}
Speed, also called as entry rate by \citeauthor{arif2009}, measures the number
of characters entered within a specific time frame. The most common metric that
measures speed is \ac{wpm}. \ac{wpm} as defined by \citeauthor{arif2009} is:

\begin{equation}
	WPM = \frac{|T| - 1}{S} \cdot 60 \cdot \frac{1}{5}
	\label{equ:wpm}
\end{equation}
\labelequations{Words per Minute}

Where, $|T|$ is the length of the text, $S$ is the time in seconds spent writing
the text. This time starts directly after the first character has been pressed,
and ends when the last letter has been entered. As such, $1$ is subtracted from
$|T|$, as the time spent to find and press the first character cannot be
accurately determined. However, some typing assessments do not subtract $1$ from
$|T|$. $60$ refers to the number of seconds in a minute and $\frac{1}{5}$
normalizes the metric for the average length of words.

Other metrics also measure speed, but they aren't as commonly used as \ac{wpm}.
These include Characters per Minute, Gestures per Second, Adjusted Words per
Minute, and Keystrokes per Second

\subsubsection{Accuracy}
Accuracy measures the number of correctly pressed characters in an input string.
Accuracy, as defined by \citeauthor{bartnik2021}, is:

\begin{equation}
	ACC = \frac{|C|}{|T|} \cdot 100\%
\end{equation}
\labelequations{Accuracy}

Where $|C|$ is the number of correct characters and $|T|$ is the length of the
text.

The inverse of accuracy is error rate, where the number of incorrectly pressed
characters is measured instead. \citeauthor{arif2009} describe 5 common error
rate metrics: Error Rate, Minimum String Distance Error Rate, Keystroke per
Character, Erroneous Keystroke Error Rate, and Total Error Rate.

\subsubsection{Limitations of the Metrics}
These metrics are all based on the inputted characters by the user. These
metrics do not take into account other aspects of keyboard typing such as
posture, hand and wrist positions, and finger placement. Consequently, these
metrics do not give a full picture of the performance of the person typing, and
they only provide a cursory view of how a person types.

\subsection{Keyboard Typing Methodology}
Keyboard typing can be accomplished in numerous ways. The main difference
between the different methodologies is the number of fingers used when typing
and how the typist navigates the keyboard to find the keys. The methodology
ranges from Hunt and Peck to Touch Typing, with variations of the two in
between.

>>>>

\documentclass[preprint,12pt,authoryear]{elsarticle}

\usepackage[utf8]{inputenc}

% Bibliograpy
\usepackage{natbib}
\usepackage{url}

% Symbols
\usepackage{gensymb}
\usepackage{amsmath}

@@

% Graphs
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{external}
\pgfplotsset{width=1\textwidth,height=0.7\textwidth,compat=1.9}

\usepackage{graphicx}
\graphicspath{ {./images/} }

\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\DeclareAcronym{roi}{
	short=ROI,
	long=Region of Interest,
	long-plural-form=Regions of Interest
}

@@

\journal{International Journal of Human-Computer Studies}

@@

\begin{frontmatter}
	\title{Development of a Finger-Key Identification Module for a Touch Typing Trainer}
	\author[inst1]{Oscar Vian L. Valles}
	\ead{oscarvianvalles@gmail.com}
	\author[inst1]{Dhong Fhel K. Gom-os\corref{cor1}}
	\ead{dkgomos@up.edu.ph}
	\cortext[cor1]{Corresponding author at: Cebu City, Cebu, Philippines}
	\affiliation[inst1]{organization={Center for Research in Intelligent Systems, Department of Computer Science, University of the Philippines Cebu},
		country={Philippines}}

	\begin{abstract}
		Typing tests list out words for the user to repeat. These tests measure metrics
		that quantify the user's ability to type. However, conventional typing metrics
		do not measure correct finger placement. This aspect of typing may affect the
		user's health. It is also a crucial aspect of keyboard typing education. As
		such, a method to identify which finger is used to press which key is
		beneficial. This paper introduces a technique to achieve this by developing a
		finger-key identification module that utilizes computer vision algorithms to
		detect the keys in a keyboard, and a ready-made machine learning solution to
		track fingers while typing. This module was successful in finger-key
		identification with an accuracy of 99.58\% in a dataset of 942 keypresses. The
		module also had an average finger-key identification time of 0.083 seconds,
		which allows the module to perform real-time finger-key identification for
		typing speeds up to 143.040 Words per Minute which is well beyond the median
		typing speed of the population. Furthermore, two metrics for correct finger
		placement were defined and were implemented in a proof-of-concept trainer: (1)
		Finger Placement Accuracy which computes the number of keys pressed using the
		correct finger in a typing test sequence, and (2) Historical Finger Placement
		Accuracy which evaluates the user's ability to press a certain key with the
		correct finger over multiple typing test sequences.
	\end{abstract}

	\begin{highlights}
		\item Automated identification of what finger was used to press a key with
		an accuracy of 99.58\% in a dataset of 942 keypresses.
		\item Finger-key identifications take an average of 0.083 seconds. This
			allows for finger-key identification for typing speeds up to 143.040 Words
			per Minute
		\item Two metrics for correct finger placement were created: Finger
		Placement Accuracy and Historical Finger Placement Accuracy
	\end{highlights}

	\begin{keyword}
		touch typing \sep keyboard typing metrics \sep touch typing trainer \sep
		computer vision \sep finger tracking
	\end{keyword}

\end{frontmatter}

\section{Introduction}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests
list out words that are then typed out. The entered keys are then compared to
check if the user has typed the expected letter \citep{bartnik2021}. At the end
of the test, the time taken is calculated, and certain metrics are given.
Examples of metrics include \ac{wpm} and Accuracy \citep{arif2009}. However,
this method of examination leaves out a crucial part of typing --— correct
finger placement.

There have been studies that show the effect of keyboard typing on the human
body. These studies have shown that keyboard typing affects our neck, shoulder,
upper limb, wrist, arms, and fingers \citep{szeto2005, baker2007digit}. In
addition, it has been shown that 22\% of computer users sustain musculoskeletal
disorders of the upper extremity. This includes the neck, shoulder, hands, and
wrists. \citep{gerr2002}

Incorrect placement of the fingers exacerbates these issues. These erroneous
finger placements may cause the following hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \citep{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions, such as in typing, may cause injuries such
as Carpal Tunnel Syndrom \citep{toosi2015} and musculoskeletal disorders of the
upper extremity (\citet{marklin1999} as cited in~\citet{baker2007}).

These issues can be prevented through keyboard typing education. Keyboard typing
has been a part of the curriculum for a long time \citep{hoot1986} and studies
have continued to this day to continue to optimize and improve methods of
teaching keyboard typing to students. By starting to teach touch typing to
students early, these students will develop the potential for higher-level
keyboard typing \citep{donica2018}.

There have been tools created to facilitate the teaching of touch typing.
Keyboarding without Tears is a web-based application and 36-week curriculum that
teaches students touch typing by following the three stages of Motor Learning
Theory \citep{kwt}. Monkeytype and TypeRacer are two keyboard typing test
websites where users type a predetermined phrase, quote, or random words, and
metrics are given after the test \citep{bartnik2021, typeracer}. Keybr is
similar to Monkeytype and Typerace, in that they also have the users type a
predetermined phrase, quote, or random words. However, this application guides
the user by using statistics to create typing lessons that are appropriate to
the current typing proficiency of the learner \citep{keybr}.

However, these typing tests only measure the ability to type the correct
character and do not check correct finger placement. While educators may be able
to weed out bad habits manually, it is impractical for them to check each
student individually.

While there have been other studies that utilize hand and finger tracking as
input methods. Examples include studies by \citet{dorf2001} which used finger
tracking as an input method in augmented environments, \citet{chiang2014} used a
Kinect to track fingers to play virtual instruments, and \citet{yousaf2014}
created a virtual keyboard that operates using finger tracking. However, these
studies used hand and finger tracking as an alternative input method, rather
than supplementing physical keyboard typing.

There is currently no solution to check for correct finger placement while
typing on a physical keyboard. Thus, there is a need for automatically
identifying which finger is used to press which key during typing --- hereby
defined as finger-key identification.

The paper aims to create a finger-key identification module that accurately
identifies what finger was used to press a key at a point in time to help
develop better typing habits and healthier typing ergonomics through the use of
computer vision algorithms. The paper also aims to create metrics to quantify
finger placement during the duration of keyboard typing.

\section{Algorithm}

\subsection{Setup}
\label{section:metho-setup}

The experimental setup and configuration controlled three key elements: the
camera, the keyboard, and the environment. The camera used was a Logitech C920.
It was placed directly above the keyboard and captured in 720p/30fps. The
keyboard used was a light-colored 60\% keyboard in a modified \ac{ansi} layout.
This type of keyboard only has the alphanumeric portion of a full-size keyboard.
The color of the surface the keyboard rested on was dark. The contrasting color
of the surface and the keyboard was chosen to improve the performance of the
algorithm in keyboard detection. The entire environment was evenly lit with a
6500k LED bulb rated at 9 watts outputting 700 lumens.

This setup was used to collect data points to measure the algorithm's ability to
perform finger-key identification in a controlled environment.

\subsection{Keyboard Detection and Key Mapping}
\label{section:metho-algo-keyboard}

A computer vision algorithm was created as a starting point for mapping the keys
of the keyboard within a video. OpenCV \citep{opencv} was used as the image
processing library for this step.

\subsubsection{Get Image Map}
The first portion of the algorithm creates an image map that is overlaid over
the detected edges of the keyboard. This image map subdivides the keyboard into
multiple \acp{roi} where each \ac{roi} represents a key.

To do so, the input frame from the camera was converted to an 8-bit grayscale
image. The frame was then denoised by using a bilateral filter. This filter
removes noise while maintaining the edges. A Sobel filter was then used for edge
detection \citep{sobel2014}. Thresholding was then performed using Otsu's
algorithm \citep{otsu} to reduce the bit-depth of the image from 8 to 1. This
reduction of bit depth improves the performance in finding contours. Contours
were found using the algorithm of \citet{contours}.

The detected contours were then simplified as the output of the algorithm
results in multiple points. To find the four points of the edges of the
keyboard, the contours were sorted and the largest one was selected. The
Douglas-Peucker algorithm for line simplification \citet{douglas-peucker} was
then performed to reduce the contour points to the minimum amount. The contour
points were then sorted clockwise, as the algorithm of \citet{contours} does not
guarantee the arrangement of the contour points. There is a need for the points
to be ordered clockwise since each point will be paired to the edges of the
image map, which were sorted in clockwise order.

The image map was then stretched over the keyboard by using a perspective warp,
where the four points of the edges of the keyboard were the end position of the
four points on the edges of the image map. Figure~\ref{fig:metho-image-map}
presents a grayscale frame and the resulting image map from that frame.

In ideal cases, the number of points would be four, with each point
corresponding to the edges of the keyboard. However, there were times when other
objects would be within the frame, or they intersected with the keyboard. This
would result in a contour that would be defined by more than four points ---
which caused the entire algorithm to fail.

\begin{figure}[h]
	\centering
	\begin{subfigure}{.50\textwidth}
		\centering
		\includegraphics[width=.995\linewidth]{grayscale.png}
		\caption{Input image that was converted to grayscale}
	\end{subfigure}%
	\begin{subfigure}{.50\textwidth}
		\centering
		\includegraphics[width=.995\linewidth]{transformed-image-map.png}
		\caption{Transformed image map}
	\end{subfigure}%
	\caption{First and last step of getting the image map}
	\label{fig:metho-image-map}
\end{figure}

\subsubsection{Get Key Contour Points}
The second portion of the algorithm pre-computes the contour points of each key
in the keyboard based on the generated virtual map and the Key-Color Values map.

The color that corresponds to each key in the virtual map is stored in the
Key-Color Value map. Each individual \ac{roi} was isolated using the color
values stored in the map. The simplified contours of each \ac{roi} were then
obtained using the same method in finding and simplifying the contour of the
keyboard.

During testing using training data, a lot of the failures in finger-key
identification were due to the tight fit of the contour to the key. The contour
on its own did not give enough buffer for the placement of the finger. This
buffer was needed as, in some instances, the finger used to press the key was
not exactly on top of the key, but rather to its side. In addition, it was also
observed that the finger tracking does not annotate the very edge of the
fingertip, but rather, it labels somewhere in the middle of the fingernail. The
algorithm accounts for this situation by adding a buffer. This was achieved by
scaling the contour points by seven pixels on each side.

After the contours of all \acp{roi} were obtained, a Key-Edge Coordinates map
was generated.

\subsection{Finger Detection and Tracking}
\label{section:metho-algo-finger}
MediaPipe Hands \citep{mediapipe} was used as the finger detection and tracking
solution. This algorithm is composed of two ML models working in conjunction to
be able to detect the different parts of the hands and track them accurately.
This algorithm outputs the positions of 21 hand and finger landmarks for each
hand within the image.

\subsection{Integration}

The two algorithms were combined to accomplish finger-key identification The
integration of the algorithm was a two-step process. The first step was to get
the Key-Edge Coordinates map shown in Section~\ref{section:metho-algo-keyboard}.
The second step runs whenever a key press has been detected. This step used the
Key-Edge Coordinates map in conjunction with the finger tracking algorithm
selected in Section~\ref{section:metho-algo-finger}.

The finger tracking data contained the pixel positions of each landmark of the
hands within the frame. The edge coordinates of the \ac{roi} of the key pressed
were obtained using the Key-Edge Coordinates map. Any fingertip landmarks that
fit within the \ac{roi} were then returned.

\section{Data Gathering}
Ten sentences were gathered from novels and texts in the public domain. These
sentences were used as typing test sequences. 3 sets of 10 videos were then
captured with each video corresponding to one typing test sequence. Each set
corresponds to three predetermined speeds of typing: slow (15wpm), average
(35wpm), and fast (80wpm). These speeds were based on test data
from~\citet{keybr}. Best effort was made to stick to the predetermined typing
speeds, however, there were discrepancies and the exact speed was not maintained
throughout the entire typing test sequence. During typing, errors were not
consciously taken into account, and errors happened naturally as part of the
typing process --- with more errors happening more frequently as the typing
speed increased.

Manual finger-key identification was performed for all key presses present in
the video. The annotated data was then divided into training and test data at a
ratio of 60:40, respectively.

\subsection{Data Inventory}

\begin{figure}[H]
	\begin{tikzpicture}
		\begin{axis}[
				xbar, xmin=0,
				bar width=3pt,
				width=0.9\textwidth,
				height=1\textwidth,
				enlarge y limits=0.02,
				xlabel=Count,
				symbolic y coords={
						Spacebar,
						Period,
						LeftShift,
						LeftAlt,
						ForwardSlash,
						Dash,
						Comma,
						Backspace,
						Z,
						Y,
						X,
						W,
						V,
						U,
						T,
						S,
						R,
						Q,
						P,
						O,
						N,
						M,
						L,
						K,
						J,
						I,
						H,
						G,
						F,
						E,
						D,
						C,
						B,
						A,
						1,
					},
				ytick=data,
				tick label style={font=\footnotesize},
				legend style={font=\footnotesize},
				label style={font=\footnotesize},
				xmajorgrids=true,
				area legend,
			]

			\addplot[
				CornflowerBlue,
				fill=CornflowerBlue
			] table[x=Value,y=Key]{test-data.txt};

			\addplot[
				Salmon,
				fill=Salmon,
			] table[x=Value,y=Key]{training-data.txt};
			\legend{Test Dataset, Training Dataset}
		\end{axis}
	\end{tikzpicture}
	\caption{Data Inventory}
	\label{fig:metho-inventory-training}
\end{figure}

Individual counts of the characters in the training and testing data set are
shown in Figure~\ref{fig:metho-inventory-training}. The letters were not
represented equally in the data set. Of note, J and Q were not in the training
dataset, and J, Q, and Z were not in the testing dataset. The inequality of
representation in both of these datasets can be attributed to the source
materials of the typing test sequences --- novels and texts. There was no
conscious decision to equally represent all characters in the data set.

\section{Training and Testing}
A script was created to test the accuracy of the module. This script opened each
video and the associated labeled file. There were three frames taken for each
video to perform step one of the module: frames ten, fifteen, and twenty. The
first frame that returned a Key-Edge Coordinates map was used. The script then
parsed the labeled file and opened the frame for each keypress. The second
algorithm of the module was then run. The result of the module was then compared
with the manual labeling of the keypress. Adjustments were then made to the
algorithm after each training pass to improve its performance.

The same script was run through the test data.
